Shapes analysis for time series.

### 2024-10-23 NEURIPS24/experiment.tex
update
improve exp

\section{Experiments}
\vspace{-1ex}
\label{section:experiments}

For conciseness, several experiments are relegated in appendix:

@@

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix: shape_classification}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/ts_ddmm_shooting.pdf}
    \caption{TS-LDDMM}
    \label{fig:ts-lddmm shooting}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/lddmm_shooting.pdf}
    \caption{LDDMM}
    \label{fig:lddmm shooting}
  \end{subfigure}
  \caption{Analysis of the two principal components (PC) related to mice' respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}, and LDDMM \Cref{fig:lddmm shooting}.
  In both cases, I) displays PC densities according to mice genotype and II) displays  deformations of the reference graph $S_0$ along each PC.}
  \label{fig:exp1}
\end{figure*}

\vspace{-1ex}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{colq cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{wt cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\Cref{fig:colq-cycle} is a \textbf{colq} respiratory cycle. \Cref*{fig: pcs-scatter} displays individual mouse reference respiratory cycle in the TS-LDDMM PC1-PC2 coordinates. \Cref{fig:wt-reference} is a referent respiratory cycle of a \textbf{wt} mouse learned with TS-LDDMM. }
  \vspace{-1em}
\end{figure*}
\vspace{-1ex}
\subsection{Interpretability: analysis of respiratory behavior in mice}
\label{section:interpretability}
\vspace{-1ex}
%We consider a dataset composed of mouse respiratory time series before and after a drug injection.
% A complete presentation of this dataset is given in \Cref{appendix:mouse_dataset}.
%  The mouse are divided in two groups depending on their genotypes : \textit{colq} and \textit{wt}.
%  We aim to study the difference of respiratory shapes according to the genotype.
%   That is why, TS-LDDMM \eqref{eq:general_optimization_problem} is applied to derive the features representations $(\mathbf{\alpha}_j)_{j\in[N]}\in (\Rset^2)^N$ related to $N$=\sam{fill} respiratory time series coming from 14 mouse (7 \textit{colq} and \textit{wt}) before drug injection.
%   Then, a Principal Components Analysis (PCA) is performed on $(\mathbf{\alpha}_j)_{j\in[N]}$

%the densities of each genotype according to each PC are displayed. In Figure B), the deformations of the reference graph $S_0$ along each PC are given. In Figure D), the graph of reference $S^j$, also called barycenter, related to each mouse, is displayed according to their coordinates on PC1 and PC3. In Figure C) et E), illustrations of respiratory cycles related to mice coming from the \textbf{wt} and \textbf{colq} group are displayed.

%\begin{figure*}[t]
%  \centering
%  \includegraphics[width=0.95\linewidth]{"./pictures/exp_1_bis.pdf"}
%  \vspace{-1.5em}
%  \caption{Analysis of the two principal components (PC) related to the mice respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}.
%  In Figure A), the densities of each genotype according to each PC are displayed. In Figure B), the deformations of the reference graph $S_0$ along each PC are given. In Figure D), the graph of reference $S^j$, also called barycenter, related to each mouse, is displayed according to their coordinates on PC1 and PC3. In Figure C) et E), illustrations of respiratory cycles related to mice coming from the \textbf{wt} and \textbf{colq} group are displayed.  }
%  \label{fig:exp_1_PCA}
%  %\vspace{-1em}
%\end{figure*}

This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in time series biomedical datasets. We consider a multivariate time series dataset that accounts for mice's nasal and thoracic airflow evolution when exposed to an irritant molecule altering respiratory functions \cite{nervo2019respiratory}.
The dataset is divided into two groups, one composed of 7 control mice (\textbf{wt}) and the other of 7 mice (\textbf{colq}) deficient in an enzyme involved in the control of respiration. For each mouse, airflows were recorded for 15 to 20 minutes before exposure to the irritant molecule and then for 35 to 40 minutes. A complete 
description of the dataset is given in the \Cref{appendix:mouse_dataset}.

\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.95\linewidth]{"./pictures/exp2.pdf"}
  \caption{Analysis of the first Principal Component (PC1) related to mice' respiratory cycles before and after exposure for TS-LDDMM. Figure A) displays PC densities according to mice genotype, Figure B) displays  deformations of the reference graph $S_0$ along PC1, and Figure C) displays respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}
\paragraph{Breathing behaviors' evolution after exposure to the irritant molecule} 
TS-LDDMM representation features are learned on a dataset that includes respiratory cycles before and after exposure.
 \Cref{fig:exp_2_PCA} displays the first principal component PC since it encodes the effect of the irritant molecule as depicted in \Cref{fig:exp_2_PCA}-C) (the exposure occurs at 20 minutes). \Cref{fig:exp_2_PCA}-B) shows that the deformation (-1.5 $\sigma_{\text{PC}}$) leads to longer respiratory cycles that include a pause between inspiration and expiration as observed in \cite{germain2023unsupervised}.
  Conjointly, \Cref{fig:exp_2_PCA}-A) shows a bimodal distribution for \textbf{wt} mice whereas \textbf{wt}' distributions before exposure were unimodal (\Cref{fig:ts-lddmm shooting}). Indeed, the irritant molecule inhibits the action of the deficient enzyme, \textbf{wt} mice strongly react to the irritant molecule, whereas \textbf{colq} mice better adapt due to their deficiency \cite{germain2023unsupervised}.
%   \vspace{-1ex}
% \paragraph{Result summary.} By comparing the shape of individual respiratory cycles, we show that TS-LDDMM features can encode genotype distinctive breathing behaviors and their evolution after exposure to the irritant molecule contrary to LDDMM \cite{glaunes2008large}. 
% \vspace{-1ex}
%Secondly, we compare breathing behaviors before and after exposure to observe the impact of the irritant molecule.
% We follow the same procedure as for before exposure, but we take $N_2=1400$ respiratory cycles extracted according 
% to the procedure CITE....
% In \Cref{fig:exp_2_PCA}, we focus on the first Principal Component (PC) since it encodes the effect of the irritant molecule as demonstrated on Figure \ref{fig:exp_2_PCA}-C) (exposure at time 20).
% We observe on \Cref{fig:exp_2_PCA}-B) that after exposure the mouse have longuer breath such that the expiration is longuer than inspiration.
%  Moreover, we deduce from Figure \ref{fig:exp_2_PCA}-A) that \textbf{colq} are more constant in their breath compared to \textbf{wt} after exposure.
%
% In Figure REF, we 
% also display the reference respiratory cycle S0 and its deformations in the principal component directions.
%  Additionally, we learn each mouse's reference respiratory cycle and represent them in the first and third PC coordinates system in Figure REF.

>>>>

\section{Experiments}
\vspace{-1ex}
\label{section:experiments}
\sam{Mettre expérience interprétation et sensibilité au bruit en appendice, améliorer l'interprétation des résultats}
For conciseness, several experiments are relegated in appendix:

@@

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix: shape_classification}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{pictures/shooting before/shooting before.pdf}
  \caption{Analysis of the two principal components (PC) related to mice ventilation before exposure with TS-LDDMM representations \textbf{(a)}, and LDDMM \textbf{(b)}. In both cases and for all PC, the left plot displays PC densities according to mice genotype and right plot displays deformations of the reference graph $\mathbf{c}_0$ along each PC.}
  \label{fig:exp1}
  \vspace{-1.5em}
\end{figure}

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{ColQ cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{WT cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\textbf{(a)} a ColQ respiratory cycle sample. \textbf{(b)} Referent respiratory cycle of individual mouse $\mathbf{c}_0^j$ in the TS-LDDMM PC1-PC2 coordinates system of $\mathbf{c}_0$. \textbf{(c)} a WT respiratory cycle sample.}
  \vspace{-1.5em}
\end{figure*}

\subsection{Interpretability: mice ventilation analysis}
\label{section:interpretability}

This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in biomedical applications. We consider a time series dataset monitoring the evolution of mice's nasal and thoracic airflow when exposed to a drug altering respiration \cite{nervo2019respiratory}.
The dataset includes recordings of 7 control mice (WT) and 7 mutant mice (ColQ) with an enzyme deficiency. The enzyme is involved in the respiration regulation, and the drug inhibits its activity. For each mouse, airflows were monitored for 15 to 20 minutes before the drug exposure and then for 35 to 40 minutes. A complete description of the dataset is given in the \Cref{appendix:mouse_dataset}.

\paragraph{Experimental protocol.} 
We considered two experimental scenarios; the first focuses on mice ventilation before exposure to explore the inter-individual and genotype-specific variabilities. The second focuses on whole recordings to analyze the evolution of mice's ventilation after drug exposure. In both cases, the baseline protocol consists of first extracting $N$ respiratory cycles from the datasets with the procedure described in \cite{germain2023unsupervised}. Then, learning the referent respiratory cycle $\mathbf{c}_0$ and the representations  of respiratory cycles $(\bm{\alpha}_0^j)_{j \in [N]}$ by solving \eqref{eq:general_optimization_problem} using TS-LDDMM. $\bm{\alpha}_0^j$ being the momentum of the initial velocity field of the geodesic encodings the diffeomorphisms mapping $\mathbf{c}_0$ to the $j^{th}$ respiratory cycle. Finally, performing a Kernel-PCA on the initial velocity fields~\eqref{eq:def_v0} belonging to $\msv$ and encoded by the pairs $(\bm{\alpha}_0^j,\mathbf{c}_0)_{j \in [N]}$. The first experiment includes $N_1 = 700$ cycles collected before exposure. The second experiment includes $N_2 = 1400$ cycles with 25\% (resp. 75\%) before (resp. after) exposure. We also performed the first experimental scenario with LDDMM representation, and \Cref{appendix: mice_exp_setting} describes the settings of both methods. Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

\paragraph{Geodesic shooting along principal component directions.}
Any principal component (PC), noted $v_0^{pc}$, from a kernel-PCA in $\msv$, is itself an initial velocity field encoded by a pair $(\mathbf{c}_0, \bm{\alpha}_0^{pc})$. PCs encode the principal axis of deformations, and it is possible to shoot along the geodesic they encode with the differential equations \eqref{eq:integration}, enabling interpretation of the main sources of deformations.

\paragraph{Mice ventilation before exposure.}
We focus on the analysis of the two first Principal Components (PC) for TS-LDDMM (\Cref{fig:exp1}a) and LDDMM (\Cref{fig:exp1}b). Looking at the geodesic shooting along PCs, \Cref{fig:exp1} shows that principal components learned with TS-LDMM lead to deformations that remain respiratory cycles. In contrast, deformations learned with LDDMM are challenging to interpret as respiratory cycles. The LDDMM velocity field kernel is a Gaussian anisotropic kernel that accounts for time and space scales; however, the entanglement of time and space dimensions in the kernel does not guarantee the graph structure, and it makes the convergence of the method complex (relative varifold loss error: TS-LDDMM: 0.06, LDDMM: 0.11).

Regarding TS-LDDMM \Cref{fig:exp1}a, its PCs refer to deformations directions carrying different physiological meanings. Indeed, the geodesic shooting along these directions indicates that PC1 accounts for variations of the total duration of a respiratory cycle, while PC2 expresses the trade-off between inspiration and expiration duration. In addition, the distribution of ColQ respiratory cycles along PC1 is wider than in WT mice, indicating that the adaptation of mutant mice to their enzyme deficiency is variable. This observation can also be seen in \Cref{fig: pcs-scatter} where a referent respiratory cycle $\mathbf{c}_0^j$ is learned by solving \eqref{eq:general_optimization_problem} for each mouse and is encoded in the (PC1,PC2) coordinate system of $\mathbf{c}_0$ by registration \eqref{eq:geodesics_shooting}. Indeed, the average respiratory cycles of ColQ mice are more spread out than WT mice's. Going back to the densities of PC1, ColQ mice distribution has a heavier tail toward negative values compared to WT mice. When shooting in the opposite direction of PC1, we can observe that the inspiration is divided into two steps. Congruently with \cite{germain2023unsupervised}, such inspirations indicate motor control difficulties due to enzyme deficiency. \Cref{fig:colq-cycle} is an example of ColQ respiratory cycle with negative PC1 coordinate.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{pictures/exp after/exp after.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with TS-LDDMM representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}

\paragraph{Mice ventilation evolution after drug exposure.} 
This experiment focuses on the first principal components learned from TS-LDDDM representations of respiratory cycles randomly sampled before and after drug exposure. \Cref{fig:exp_2_PCA}a illustrates the geodesic shootings along PC1. Again, PC1 accounts for variations in respiratory cycle duration, but more importantly, it can be observed on the deformation at -1.5 $\sigma_{\text{PC}}$ the apparition of a long pause after inspiration. Congruently, \Cref{fig:exp_2_PCA}c indicates that pauses appear after drug exposure as cycles with negative PC1 values mainly occur after 20 minutes and present more variability along PC3. In addition, \Cref{fig:exp_2_PCA}b shows a bimodal distribution for WT mice with one of the peaks in the negative values. This peak was not observed in the previous experiment \Cref{fig:exp1}a. It indicates that pauses after inspiration are prevalent in WT mice after drug exposure. On the other hand, the distributions of ColQ mice's respiratory cycles along PC1 in both experiments are similar and account for the same deformation, suggesting that ColQ mice weakly react to the drug exposure as they already adapt their enzyme deficiency.

\paragraph{Experiment Conclusion.}
Analyzing mice ventilation with TS-LDDMM representation highlights the method's ability to create meaningful interaction between experts and the data. Indeed, combining statistical and visual results shows that main deformations carry physiological meaning, enabling the characterization of some mice genotypes and the effects of drug exposure.

\iffalse
\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.95\linewidth]{pictures/exp after/exp after.pdf}
  \caption{Analysis of the first Principal Component (PC1) related to mice ventilation before and after exposure with TS-LDDMM representations. \textbf{(a)} displays PC densities per mice genotype, \textbf{(b)} illustrates deformations of the reference respiratory cycle $\mmathbf{c}_0$ along PC1, and \textbf{(c)} displays all respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCAold}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}
\paragraph{Breathing behaviors' evolution after exposure to the irritant molecule} 
TS-LDDMM representation features are learned on a dataset that includes respiratory cycles before and after exposure.
 \Cref{fig:exp_2_PCA} displays the first principal component PC since it encodes the effect of the irritant molecule as depicted in \Cref{fig:exp_2_PCA}-C) (the exposure occurs at 20 minutes). \Cref{fig:exp_2_PCA}-B) shows that the deformation (-1.5 $\sigma_{\text{PC}}$) leads to longer respiratory cycles that include a pause between inspiration and expiration as observed in \cite{germain2023unsupervised}.
  Conjointly, \Cref{fig:exp_2_PCA}-A) shows a bimodal distribution for \textbf{wt} mice whereas \textbf{wt}' distributions before exposure were unimodal (\Cref{fig:ts-lddmm shooting}). Indeed, the irritant molecule inhibits the action of the deficient enzyme, \textbf{wt} mice strongly react to the irritant molecule, whereas \textbf{colq} mice better adapt due to their deficiency \cite{germain2023unsupervised}.
%   \vspace{-1ex}
% \paragraph{Result summary.} By comparing the shape of individual respiratory cycles, we show that TS-LDDMM features can encode genotype distinctive breathing behaviors and their evolution after exposure to the irritant molecule contrary to LDDMM \cite{glaunes2008large}. 
% \vspace{-1ex}
%Secondly, we compare breathing behaviors before and after exposure to observe the impact of the irritant molecule.
% We follow the same procedure as for before exposure, but we take $N_2=1400$ respiratory cycles extracted according 
% to the procedure CITE....
% In \Cref{fig:exp_2_PCA}, we focus on the first Principal Component (PC) since it encodes the effect of the irritant molecule as demonstrated on Figure \ref{fig:exp_2_PCA}-C) (exposure at time 20).
% We observe on \Cref{fig:exp_2_PCA}-B) that after exposure the mouse have longuer breath such that the expiration is longuer than inspiration.
%  Moreover, we deduce from Figure \ref{fig:exp_2_PCA}-A) that \textbf{colq} are more constant in their breath compared to \textbf{wt} after exposure.
%
% In Figure REF, we 
% also display the reference respiratory cycle S0 and its deformations in the principal component directions.
%  Additionally, we learn each mouse's reference respiratory cycle and represent them in the first and third PC coordinates system in Figure REF.

\fi
